---
title: "R Notebook"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

If you opened this rmd directly, the file paths will not work, be sure
you opened the .rproj file and then opened the this rmd within rstudio

Leave coverages in current folders or code won't work!

```{r echo = FALSE}
library(tidyverse)
library(tmap)
library(terra) #the raster package is going away, switching to terra
library(raster)
library(whitebox)
library(RSAGA)
library(sf)
library(qgisprocess)
library(caret)

algos <- qgis_algorithms()
#install_whitebox()
wbt_init()
```

Whitebox tools manual:
<https://www.whiteboxgeo.com/manual/wbt_book/intro.html> RSAGA:
<https://cran.r-project.org/web/packages/RSAGA/RSAGA.pdf>

Prep DEM for hydro related metrics and make 5m regular dem

Feature preserving smooth Fill single cell pits Breach larger
depressions

```{r}
#originaldem <- "knb-lter-hbr.211.2/dem1m.tif"
#hydro_prepped_orig_dem <- "knb-lter-hbr.211.2/hydem1mlpns.tif"

data_dir <- "ConePond/"
originaldem <- "cnp1mdem.tif"
hydro_prepped_orig_dem <- "cnp1mdem.tif"

#data_dir <- "BlackPond/" #with a / after ie. directory/
#originaldem <- "bk1mdem.tif"
#hydro_prepped_orig_dem <- "bk1mdem.tif"
#epsg_orig <- "epsg:26919"
#
#data_dir <- "HBValley/" #with a / after ie. directory/
#originaldem <- "knb-lter-hbr.211.2/dem1m.tif"
#hydro_prepped_orig_dem <- "knb-lter-hbr.211.2/hydem1mlpns.tif"
epsg_orig <- "epsg:26919"


#Aggregate 1 m DEM to 5 m
wbt_aggregate_raster(
  input = paste0(data_dir, originaldem),
  output = paste0(data_dir, "dem5m_noproj.tif"),
  agg_factor = 5
  )

#aggregate strips the projection, this adds it back in
temp <- rast(paste0(data_dir, "dem5m_noproj.tif"))
crs(temp) <- epsg_orig
writeRaster(temp, 
           paste0(data_dir, "dem5m.tif"),
           overwrite = TRUE)

#this strips the projection
#must add the projection back after
#calculating downslope index
#downslope index tool does not like 
#rasters output by terra functions
wbt_aggregate_raster(
  input = paste0(data_dir, hydro_prepped_orig_dem),
  output = paste0(data_dir, "hydem5m.tif"),
  agg_factor = 5
  )

#smooth 5m DEM
wbt_feature_preserving_smoothing(
  dem = paste0(data_dir, "hydem5m.tif"),
  output = paste0(data_dir, "hydem5m_smooth.tif"),
  filter = 11,
  norm_diff = 15,
  num_iter = 3,
  max_diff = 0.5)

#breach depressions
wbt_breach_depressions_least_cost(
  dem = paste0(data_dir, "hydem5m_smooth.tif"),
  output = paste0(data_dir, "hydem5m_smooth_breached.tif"),
  dist = 100,
  flat_increment = 0.01)

#fill remaining depressions there is 
#a bug in the breach depressions least cost
#function that leaves some depressions, 
#this fixes that
wbt_fill_depressions_wang_and_liu(
  dem = paste0(data_dir, "hydem5m_smooth_breached.tif"),
  output = paste0(data_dir, "hydem5m_smooth_breached_filled.tif")
)

```

Code used to check on DEMs generated above
```{r}
wbt_hillshade(
  dem = paste0(data_dir, "hydem5m_smooth_breached_filled.tif"),
  output = paste0(data_dir, "hydem5m_prepped_hillshade.tif"))

hillshade <- rast(paste0(data_dir, "hydem5m_prepped_hillshade.tif"))

tmap_mode("plot")
tm_shape(hillshade)+
  tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
  tm_scale_bar()
```

Original topo metrics (Gillin et al 2015) TPI 100 TWId EDb Uaab 

New metrics moving forward in Whitebox or RSAGA Max slope (1m) (use 1m DEM)

TPI 20, 100, 500 (SAGA, DevFromMeanElev in WBT?) (5m but not smoothed)
Multi-resolution valley bottom (MRVB; SAGA; 5m not smoothed) TWId
(GRASS, raster calculator; 5m, hydro prepped) Feature preserving filter

DEM (GRASS) Ln(UAA/tan(downslope index 5m distance))? SAGA \#downslope
unsaturated flowpath with WBT free trial (1m, 5m, smooth/not)

TPI
<https://www.whiteboxgeo.com/manual/wbt_book/available_tools/geomorphometric_analysis.html?highlight=topographic%20position#relativetopographicposition>

```{r}

#qgis_show_help("saga:slopeaspectcurvature")
dem <- rast(paste0(data_dir, "hydem5m.tif"))

#slope in radians
qgis_run_algorithm(algorithm = "saga:slopeaspectcurvature",
                  ELEVATION = dem,
                  METHOD = 0,
                  UNIT_SLOPE = 2,
                  SLOPE = paste0(data_dir, "slope_per_qgis_5m")
                   )

qgis_slope <- rast(paste0(data_dir, "slope_per_qgis_5m.sdat"))
crs(qgis_slope) <- "epsg:26919"
writeRaster(qgis_slope, paste0(data_dir, "slope_per_qgis.tif"), overwrite = TRUE)

#slope in percent
qgis_run_algorithm(algorithm = "saga:slopeaspectcurvature",
                  ELEVATION = dem,
                  METHOD = 0,
                  UNIT_SLOPE = 0,
                  SLOPE = paste0(data_dir, "slope_rad_qgis_5m")
                   )

qgis_slope <- rast(paste0(data_dir, "slope_rad_qgis_5m.sdat"))
crs(qgis_slope) <- "epsg:26919"
writeRaster(qgis_slope, paste0(data_dir, "slope_rad_qgis.tif"), overwrite = TRUE)

#log md inf (Seibert & McGlynn 07) flow accumulation
#1.65 seconds
wbt_md_inf_flow_accumulation(
  dem = paste0(data_dir, "hydem5m_smooth_breached_filled.tif"),
  output = paste0(data_dir, "hydem5m_mdinf_fa.tif"),
  out_type = "specific contributing area"
)

#tangent downslope index
wbt_downslope_index(
  dem = paste0(data_dir, "hydem5m_smooth_breached_filled.tif"),
  output= paste0(data_dir, "hydem5m_downslope_index.tif"),
  drop = 5,
  out_type = "tangent"
)

#read flow accumulation and downslope index for raster calc operation
mdinf <- rast(paste0(data_dir, "hydem5m_mdinf_fa.tif"))
tan_ds_index <- rast(paste0(data_dir, "hydem5m_downslope_index.tif"))

#divide natural log flow accumulation by tangent of downslope index
# + 0.00001 to eliminate inf values bc of division by zero
TWId <- base::log(mdinf / (tan_ds_index + 0.000001)) #log10?

crs(TWId) <- epsg_orig

#output TWId to file
writeRaster(TWId, paste0(data_dir, "hydem5m_TWId.tif"), overwrite = TRUE)
```


```{r}
#qgis_show_help("saga:topographicpositionindextpi")
dem <- rast(paste0(data_dir, "hydem5m.tif"))

sagatpi <- function(dem, data_dir, radius){
  qgis_run_algorithm(algorithm = "saga:topographicpositionindextpi",
                  STANDARD = 0,
                  DEM = dem,
                  RADIUS_MIN = radius,
                  RADIUS_MAX = radius,
                  TPI = paste0(data_dir, "tpi", radius, "saga")
                   )
  
  tpisaga <- rast(paste0(data_dir, "tpi", radius, "saga.sdat"))
  
  crs(tpisaga) <- "epsg:26919"
  
  writeRaster(tpisaga, 
              paste0(data_dir, "tpi", radius, "saga.tif"), 
              overwrite = TRUE)
}

sagatpi(dem, data_dir, radius = 15)
sagatpi(dem, data_dir, radius = 20)
sagatpi(dem, data_dir, radius = 100)
sagatpi(dem, data_dir, radius = 200)
sagatpi(dem, data_dir, radius = 250)
sagatpi(dem, data_dir, radius = 500)
sagatpi(dem, data_dir, radius = 1000)
sagatpi(dem, data_dir, radius = 1500)
sagatpi(dem, data_dir, radius = 2000)
sagatpi(dem, data_dir, radius = 2500)


```
Run BOSS model to predict bedrock
```{r}
source("boss_fxn.R")

boss(dir = data_dir,
     maxslope = raster(paste0(data_dir,"slope_rad_qgis.tif")),
     tpi15 = raster(paste0(data_dir,"tpi15saga.tif")),
     tpi200 = raster(paste0(data_dir,"tpi200saga.tif")),
     model = "bossmodel2024-02-07.rds")

plot(raster(paste0(data_dir,"bossBRprediction_sagatpi.tif")))
```

Euclidean distance from bedrock
```{r}
#euclidian distance from bedrock
boss01 <- raster(paste0(data_dir,"bossBRprediction.tif"))
#boss01[boss01 == 1] <- 0
#boss01[boss01 == 2] <- 1 #BR == 2 means bedrock in current prediction model

writeRaster(boss01, paste0(data_dir, "bossBRprediction.tif"), overwrite = TRUE)

wbt_euclidean_distance(
      input = paste0(data_dir,"bossBRprediction.tif"),
      output = paste0(data_dir, "EDb.tif"))

```

Make UAAb bedrock weighted UAA using QGIS
```{r}
#prep BR weight grid by making BR values 10000
BR <- raster(paste0(data_dir,"bossBRprediction.tif"))
BR[BR == 1] <- 100000 #BR == 2 means bedrock in current prediction model
BR[BR == 0] <- 0
writeRaster(BR, paste0(data_dir, "BRweights.tif"), overwrite = TRUE)

#use QGIS to make UAAb
weights <- raster(paste0(data_dir,"BRweights.tif"))

qgis_run_algorithm(algorithm = "sagang:catchmentarearecursive",
                  FLOW_UNIT = 0,
                  ELEVATION = dem,
                  METHOD = 3,
                  NO_NEGATIVES = 1,
                  ACCU_TARGET = dem,
                  FLOW = paste0(data_dir, "catcharea")
                   )

catcharea <- rast(paste0(data_dir, "catcharea.sdat"))
crs(catcharea) <- "epsg:26919"
writeRaster(catcharea, paste0(data_dir, "catcharea.tif"), overwrite = TRUE)

#weighted
#qgis_show_help("saga:catchmentarearecursive")
qgis_run_algorithm(algorithm = "sagang:catchmentarearecursive",
                  ELEVATION = dem,
                  FLOW_UNIT = 0,
                  METHOD = 3,
                  WEIGHTS = weights,
                  NO_NEGATIVES = 1,
                  ACCU_TARGET = dem,
                  FLOW = paste0(data_dir, "catcharea_br")
                   )

catcharea_br <- rast(paste0(data_dir, "catcharea_br.sdat"))
crs(catcharea_br) <- "epsg:26919"
writeRaster(catcharea_br, paste0(data_dir, "catcharea_br.tif"), overwrite = TRUE)

diff <- catcharea_br/catcharea
normdiff <- (diff - minmax(diff)[1]) / (minmax(diff)[2] - minmax(diff)[1])
writeRaster(normdiff, paste0(data_dir, "uaab_norm2.tif"), overwrite = TRUE)

####
###
###
##TEST WITH ORIGINAL BR

##prep BR weight grid by making BR values 10000
#BR <- raster("Original_Metrics/bedrock/bedrock01.tif")
#BR[BR == 1] <- 100000 #BR == 2 means bedrock in current prediction model
#BR[BR == 0] <- 0
#writeRaster(BR, paste0(data_dir, "BRweightsOG.tif"), overwrite = TRUE)
#
##use QGIS to make UAAb
#weights <- raster(paste0(data_dir,"BRweightsOG.tif"))
#
#qgis_run_algorithm(algorithm = "saga:catchmentarearecursive",
#                  FLOW_UNIT = 0,
#                  ELEVATION = dem,
#                  METHOD = 3,
#                  NO_NEGATIVES = 1,
#                  FLOW = paste0(data_dir, "catcharea")
#                   )
#
#catcharea <- rast(paste0(data_dir, "catcharea.sdat"))
#crs(catcharea) <- "epsg:26919"
#writeRaster(catcharea, paste0(data_dir, "catcharea.tif"), overwrite = TRUE)
#
##weighted
##qgis_show_help("saga:catchmentarearecursive")
#qgis_run_algorithm(algorithm = "saga:catchmentarearecursive",
#                  ELEVATION = dem,
#                  FLOW_UNIT = 0,
#                  METHOD = 3,
#                  WEIGHTS = weights,
#                  NO_NEGATIVES = 1,
#                  FLOW = paste0(data_dir, "catcharea_br")
#                   )
#
#catcharea_br <- rast(paste0(data_dir, "catcharea_br.sdat"))
#crs(catcharea_br) <- "epsg:26919"
#writeRaster(catcharea_br, paste0(data_dir, "catcharea_br_.tif"), overwrite = TRUE)
#
#diff <- catcharea_br/catcharea
#normdiff <- (diff - minmax(diff)[1]) / (minmax(diff)[2] - minmax(diff)[1])
#writeRaster(normdiff, paste0(data_dir, "uaab_norm2.tif"), overwrite = TRUE)
#
#
```
MRVBF
```{r}
###MRVBF in QGIS package
dem <- raster(paste0(data_dir, "hydem5m.tif"))

#qgis_show_help("saga:multiresolutionindexofvalleybottomflatnessmrvbf")
qgis_run_algorithm(algorithm = "saga:multiresolutionindexofvalleybottomflatnessmrvbf",
                  DEM = dem,
                  MRVBF = paste0(data_dir, "mrvbf_qgis")
                   )

mrvbf <- rast(paste0(data_dir, "mrvbf_qgis.sdat"))
crs(mrvbf) <- "epsg:26919"
writeRaster(mrvbf, paste0(data_dir, "mrvbf.tif"), overwrite = TRUE)
```
run HPU Model
```{r}
source("PSM_HPU_FXN.R")

HPU_predict(
    data_dir = data_dir,
    hbmaxslope = raster(paste0(data_dir,"slope_per_qgis.tif")),
    hbuaab     = raster(paste0(data_dir,"uaab_norm2.tif")),
    twid       = raster(paste0(data_dir,"hydem5m_TWId.tif")), 
    mrvbf      = raster(paste0(data_dir,"mrvbf.tif")),
    tpi20      = raster(paste0(data_dir,"tpi20saga.tif")),
    tpi100     = raster(paste0(data_dir,"tpi100saga.tif")),
    tpi200     = raster(paste0(data_dir,"tpi200saga.tif")),
    EDb        = raster(paste0(data_dir, "EDb.tif")),
    model      = readRDS("HBValley/model2024-01-31.rds"))
```



